Soss

### 2019-12-31 paper/paper.tex
edits

\begin{document}

@@

\begin{abstract}
We present Soss, a probabilistic programming language embedded in the Julia language. Soss represents statistical models in terms of abstract syntax trees, and uses staged compilation for on-demand generation of ``inference primitives'' (random sampling, log-density, etc) without requiring casual users to worry about such details.

@@

Here's some example code. Lots of dead space on the right, maybe want to try to keep code in a multi-column figure

@@

    \caption{A plate diagram (a) and Soss encoding (b) for a two-way analysis of variance model.}
    \label{}
\end{figure}

@@

@@

\[
\begin{aligned}
    \ell = 
       &- 3 \log{\left(2 \right)} 
        - 5 \log{\left(\pi \right)} 
    \\ &- σ^{2} 
        - 4 \log{\left(λ \right)} 
        - 2 \log{\left(λ^{2} + 1 \right)} 
        - \frac{σ_I^{2}}{λ^{2}} 
        - \frac{σ_J^{2}}{λ^{2}}
    \\ &+ \sum_{i=1}^{I} \left(- \log{\left(σ_I \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{{α}_{i}^{2}}{2 σ_I^{2}}\right) 
    \\ &+ \sum_{j=1}^{J} \left(- \log{\left(σ_J \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{{β}_{j}^{2}}{2 σ_J^{2}}\right) 
    \\ &+ \sum_{\substack{1 \leq i \leq I \\ 1 \leq j \leq J}} \left(- \log{\left(σ \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{\left({y}_{ij} - {\hat{y}}_{ij}\right)^{2}}{2 σ^{2}}\right) 
\end{aligned}
\]

\[
    \begin{aligned}
    \ell =
    &- 7.8 - 0.92 I - 0.92 J - 0.92 I J 
    \\ &- 4 \log{λ } - 2 \log{\left(λ^{2} + 1 \right)} 
        - σ^{2} 
        - I J \log{σ } 
    \\ &- I \log{σ_I } - J \log{σ_J } - \frac{σ_I^{2}}{λ^{2}}  - \frac{σ_J^{2}}{λ^{2}}
    \\ &- \frac{1}{2 σ_I^{2}} \sum_{i=1}^{I} {α}_{i}^{2}
        - \frac{1}{2 σ_J^{2}} \sum_{j=1}^{J} {β}_{j}^{2}
        - \frac{1}{2 σ^{2}} \sum_{\substack{1 \leq i \leq I\\1 \leq j \leq J}} \left({y}_{ij} - {\hat{y}}_{ij}\right)^{2}
    \end{aligned}
\]

@@

Authors may provide appendices to accompany their extended abstract submissions. Please note however that reviewers will not be required to comment on material in these appendices.
\end{document}

>>>>

\newmintinline[jl]{julia}{}
\begin{document}

@@

\begin{abstract}
We present Soss, a declarative probabilistic programming language embedded in the Julia language. Soss represents statistical models in terms of abstract syntax trees, and uses staged compilation for on-demand generation of ``inference primitives'' (random sampling, log-density, etc) without requiring casual users to worry about such details.

@@

@@

    \caption{A plate diagram (a) and Soss encoding (b) for a two-way analysis of variance model. Note the specification that $y$ will later be observed is not part of the model, but is instead given at inference time.}
    \label{fig:model}
\end{figure}

Figure~\ref{fig:model} shows a plate diagram and Soss implementation of a simplified two-way ANOVA model. The model $m$ has \emph{arguments} $I$ and $J$. Each line of a model is either a \emph{sample statement} (\jl#v ~ rhs#) or an \emph{assign statement} (\jl#v = rhs#). In either case, \jl#v# must be a valid Julia variable name, and \jl#rhs# must be a valid Julia expression. For a sample statement, \jl#rhs# should also be ``distribution-like'', supporting \jl#logpdf# and/or \jl#rand# methods (depending on the inference algorithm to be used).

\section{Inference Primitives}

A given inference algorithm can be expressed in terms of a collection of functions on a model. For example, we may need to evaluate the log-density or its gradient, map continuous variables into $\mathbb{R}^n$, or draw random samples. In Soss, these \emph{inference primitives} are

\begin{figure}

    \begin{minted}{julia}
        julia> truth = rand(m(I=2,J=3)); pairs(truth)
        pairs(::NamedTuple) with 10 entries:
          :I    => 2
          :J    => 3
          :yhat => [0.0996722 0.720247 -0.706772; 0.100505 0.72108 -0.705939]
          :σ    => 0.297288
          :λ    => 0.523386
          :σJ   => 0.312794
          :σI   => 0.0054669
          :β    => [0.0973136, 0.717889, -0.70913]
          :α    => [0.00235854, 0.00319107]
          :y    => [0.386041 0.564963 -0.72177; 0.236898 0.842491 -0.912154]
    \end{minted}

\end{figure}

@@

\begin{minted}{julia}
julia> markovBlanket(m, :α)
    @model (I, σI, β, σ, J) begin
            α ~ Normal(0, σI) |> iid(I)
            yhat = α .+ β'
            y ~ For(I, J) do i, j
                    Normal(yhat[i, j], σ)
                end
        end
\end{minted}

@@

@@

\section{Symbolic Simplification}
Before simplification:
\[
\begin{aligned}
    \ell = 
       &- 3 \log{\left(2 \right)} 
        - 5 \log{\left(\pi \right)} 
    \\ &- σ^{2} 
        - 4 \log{\left(λ \right)} 
        - 2 \log{\left(λ^{2} + 1 \right)} 
        - \frac{σ_I^{2}}{λ^{2}} 
        - \frac{σ_J^{2}}{λ^{2}}
    \\ &+ \sum_{i=1}^{I} \left(- \log{\left(σ_I \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{{α}_{i}^{2}}{2 σ_I^{2}}\right) 
    \\ &+ \sum_{j=1}^{J} \left(- \log{\left(σ_J \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{{β}_{j}^{2}}{2 σ_J^{2}}\right) 
    \\ &+ \sum_{\substack{1 \leq i \leq I \\ 1 \leq j \leq J}} \left(- \log{\left(σ \right)} - \frac{\log{\pi }}{2} - \frac{\log{2 }}{2} - \frac{\left({y}_{ij} - {\hat{y}}_{ij}\right)^{2}}{2 σ^{2}}\right) 
\end{aligned}
\]

After simplification:
\[
    \begin{aligned}
    \ell =
    &- 7.8 - 0.92 I - 0.92 J - 0.92 I J 
    \\ &- 4 \log{λ } - 2 \log{\left(λ^{2} + 1 \right)} 
        - σ^{2} 
        - I J \log{σ } 
    \\ &- I \log{σ_I } - J \log{σ_J } - \frac{σ_I^{2}}{λ^{2}}  - \frac{σ_J^{2}}{λ^{2}}
    \\ &- \frac{1}{2 σ_I^{2}} \sum_{i=1}^{I} {α}_{i}^{2}
        - \frac{1}{2 σ_J^{2}} \sum_{j=1}^{J} {β}_{j}^{2}
        - \frac{1}{2 σ^{2}} \sum_{\substack{1 \leq i \leq I\\1 \leq j \leq J}} \left({y}_{ij} - {\hat{y}}_{ij}\right)^{2}
    \end{aligned}
\]

\end{document}
